Mamba vs SOTA for Vision Tasks\\ \thanks{$\dagger$ Equal contributions}

### 2024-04-25 paper/main.tex
add

\section{Application: Image Segmentation}

@@

\begin{thebibliography}{00}
\bibitem{brown2020language} T. B. Brown et al., "Language Models are Few-Shot Learners," arXiv:2005.14165 [cs.CL], 2020.
\bibitem{dosovitskiy2021image} A. Dosovitskiy et al., "An Image is Worth 16x16 Words: Transformers for Image Recognition at Scale," arXiv:2010.11929 [cs.CV], 2021.
\bibitem{warden2019tinyml} P. Warden and D. Situnayake, "TinyML: Machine Learning with TensorFlow Lite on Arduino and Ultra-Low-Power Microcontrollers," O'Reilly Media, Inc., 2019.
\bibitem{banbury2021benchmarking} C. R. Banbury et al., "MLPerf Tiny Benchmark," arXiv:2106.07597 [cs.LG], 2021.
\bibitem{ronneberger2015unet} O. Ronneberger, P. Fischer, and T. Brox, "U-Net: Convolutional Networks for Biomedical Image Segmentation," arXiv:1505.04597 [cs.CV], 2015.
\bibitem{he2016deep} K. He, X. Zhang, S. Ren, and J. Sun, "Deep Residual Learning for Image Recognition," arXiv:1512.03385 [cs.CV], 2016.
\bibitem{vaswani2017attention} A. Vaswani et al., "Attention Is All You Need," arXiv:1706.03762 [cs.CL], 2017.
\bibitem{choromanski2020rethinking} K. Choromanski et al., "Rethinking Attention with Performers," arXiv:2009.14794 [cs.LG], 2020.
\bibitem{gu2022efficiently} A. Gu, K. Goel, and C. Ré, "Efficiently Modeling Long Sequences with Structured State Spaces," arXiv:2111.00396 [cs.LG], 2022.
\bibitem{gu2023mamba} A. Gu and T. Dao, "Mamba: Linear-Time Sequence Modeling with Selective State Spaces," arXiv:2312.00752 [cs.LG], 2023.
\bibitem{krizhevsky2012imagenet} - ImageNet Classification with Deep Convolutional Neural Networks
\bibitem{simonyan2014very} - Very Deep Convolutional Networks for Large-Scale Image Recognition
\bibitem{szegedy2015going} - Going Deeper with Convolutions
\bibitem{he2016deep} - Deep Residual Learning for Image Recognition
\bibitem{howard2017mobilenets} - MobileNets: Efficient Convolutional Neural Networks for Mobile Vision Applications
\bibitem{tan2019efficientnet} - EfficientNet: Rethinking Model Scaling for Convolutional Neural Networks
\bibitem{radosavovic2020designing} - Designing Network Design Spaces
\bibitem{dosovitskiy2021image} - An Image is Worth 16x16 Words: Transformers for Image Recognition at Scale
\bibitem{xie2021segformer} - SegFormer: Simple and Efficient Design for Semantic Segmentation with Transformers
\bibitem{choromanski2020rethinking} - Rethinking Attention with Performers
\bibitem{cheng2017survey} - A Survey of Model Compression and Acceleration for Deep Neural Networks
\bibitem{hinton2015distilling} - Distilling the Knowledge in a Neural Network
\bibitem{gu2022efficiently} A. Gu, K. Goel, and C. Ré, "Efficiently Modeling Long Sequences with Structured State Spaces," arXiv preprint arXiv:2111.00396, 2022.
\bibitem{gu2020hippo} A. Gu, T. Dao, S. Ermon, A. Rudra, and C. Ré, "HiPPO: Recurrent Memory with Optimal Polynomial Projections," Advances in Neural Information Processing Systems, vol. 33, pp. 1474-1487, 2020.
\bibitem{gu2023mamba} A. Gu and T. Dao, "Mamba: Linear-Time Sequence Modeling with State Spaces," arXiv preprint arXiv:2302.01327, 2023.
\end{thebibliography}

>>>>

\section{Application: Image Segmentation}
Here we investigate Mamba-based architectures with regards to the task of image segmentation.

\section{Input Data}\label{sec2}
We shall be using a Brain MRI segmentation dataset found in \cite{dataset}. This dataset has been used in several papers regarding classification of the shape and severity of tumors, and provides an adequate dataset for research purposes in the field of medical image analysis, particularly in the area of brain MRI segmentation. Researchers have utilized this dataset in various studies focusing on the classification of tumor shapes and the assessment of tumor severity. With its comprehensive collection of brain MRI scans, along with corresponding segmentation masks, the dataset offers valuable resources for developing and evaluating algorithms aimed at automating tumor detection and analysis.

The dataset contains brain MRI images along with their segmentation masks for 110 patients. There were a total of 3143 train images, 393 validation images, and 393 test images. The distribution of the tumor and non-tumor images is shown in Figure \ref{fig:datastats}.

\section{Output Data}\label{sec3}
The output of this model shall be segmentation masks for new, never before seen images.
\begin{figure}[!t]
    \centering
    \includegraphics[width=\columnwidth]{imgs/data_stats.png}
    \caption{Distributions of tumor and non-tumor images in the train, validation, and test datasets.}
    \label{fig:datastats}
\end{figure}

\section{Compared Models}
Prior to an analysis of how the Pyramidal U-Mamba compares, we shall explain what models we chose for comparison and why.

For this our analysis, we compare four models against our own developed ones. The models are U-Net\cite{ronneberger2015unet}, ResNet18 \cite{DBLP:journals/corr/HeZRS15}, ResNet50, U-Mamba\cite{U-Mamba}, and SegViT\cite{dosovitskiy2021image}. Below we explain our choices.
\subsection{U-Net}
The core idea behind U-Net is to complement a traditional contracting network with successive layers that replace pooling operations with upsampling operators. These upsampling layers aim to increase the resolution of the output. Subsequent convolutional layers can then learn to assemble a precise output based on this high-resolution information \cite{ronneberger2015unet}.
A key modification in the U-Net architecture is the inclusion of a large number of feature channels in the upsampling part. This allows the network to effectively propagate contextual information to higher resolution layers. As a result, the expansive path of the network becomes more or less symmetric to the contracting part, leading to a distinctive U-shaped architecture. To predict pixels at the border regions of the image, the missing context is extrapolated by mirroring the input image. This tiling strategy is crucial for applying the network to large images, as it circumvents resolution limitations imposed by GPU memory constraints \cite{ronneberger2015unet}. Due to its encoder-decoder architecture, skip connections, multi-scale feature extraction, and its efficiency, U-Net has been used in many denoising and diffusion models. For instance, DDPMs (Denoising Diffusion Probabilistic Models) use an architecture similar to U-Net for denoising and sample generation \cite{DBLP:journals/corr/abs-2006-11239} while the popular Stable Diffusion architecture uses a U-Net based architecture for the diffusion process \cite{rombach2021highresolution}. We have chosen to compare our novel segmentation architecture against the U-Net architecture due to its well-established reputation and widespread adoption in the field of medical image segmentation.

\subsection{ResNet}
We have also selected ResNet18 and ResNet50 as additional benchmarks. These architectures, introduced by He et al. in their seminal work "Deep Residual Learning for Image Recognition" \cite{DBLP:journals/corr/HeZRS15}, have revolutionized the field of deep learning by addressing the problem of vanishing gradients in deep neural networks. The key innovation in ResNets is the introduction of residual connections, which allow the network to learn residual functions with reference to the input layer, thereby facilitating the training of much deeper networks. ResNet18 and ResNet50, with 18 and 50 layers respectively, have been widely adopted in various computer vision tasks, including image classification, object detection, and segmentation. These models have demonstrated exceptional performance and generalization ability across diverse datasets. By comparing our proposed architecture against ResNet18 and ResNet50, we aim to assess its effectiveness in relation to these well-established and highly influential architectures. This comparison will provide valuable insights into the capabilities of our model and its potential to advance the state-of-the-art in image segmentation tasks.
\subsection{U-Mamba}
As our model takes much inspiration from the recently developed U-Mamba design \cite{U-Mamba}, we also choose to incorporate it in our comparison. U-Mamba addresses these limitations by introducing a novel hybrid CNN-SSM block that leverages the strengths of both architectures. The convolutional layers in the block are responsible for local feature extraction, while the State Space Sequence Models (SSMs) \cite{gu2022efficiently}, a new family of deep sequence models, are known for their strong capability in handling long sequences and capturing long-range dependencies. By integrating these two components, U-Mamba achieves a balance between local and global information processing, enabling it to effectively handle long-range dependencies in biomedical image segmentation tasks. Moreover, U-Mamba incorporates a self-configuring mechanism that allows it to automatically adapt to various datasets without manual intervention, enhancing its versatility and usability.
\subsection{SegFormer}
SegFormer \cite{xie2021segformer} is a transformer-based semantic segmentation model that employs a hierarchical structure and a novel attention mechanism called Efficient Self-Attention (ESA). The SegFormer architecture consists of a transformer encoder for capturing long-range dependencies and a lightweight All-MLP decoder for generating high-resolution segmentation masks. The ESA module reduces the computational complexity of self-attention by performing attention operations in a local window and aggregating global information through a depth-wise convolution. This allows SegFormer to efficiently process high-resolution images while capturing both local and global context. Moreover, SegFormer introduces a position-sensitive embedding scheme that encodes both spatial and channel-wise information, enhancing the model's ability to capture fine-grained details. By comparing U-Mamba against SegFormer, we aim to evaluate the effectiveness of our hybrid CNN-SSM approach in relation to the transformer-based segmentation mechanism employed by SegFormer. This comparison will provide insights into the strengths and weaknesses of both architectures and their ability to handle complex biomedical image segmentation tasks. Furthermore, it will help us understand the potential of transformer-based approaches and hybrid approaches in advancing the state-of-the-art in biomedical image segmentation.

\subsection{UltraLight VM-UNet}
The UltraLight VM-UNet \cite{ultralightvmunet} is a lightweight neural network architecture designed for skin lesion segmentation tasks. It is built upon the Vision Mamba module, which is a state-space model (SSM) that can efficiently handle long-range dependencies in sequences, making it well-suited for image segmentation tasks.

The key innovation of the UltraLight VM-UNet is the proposed Parallel Vision Mamba Layer (PVM Layer), which processes deep features in parallel using multiple Vision Mamba blocks. Specifically, the input feature map is split into multiple sub-feature maps, each processed by a separate Vision Mamba block with a reduced channel count. This parallel processing approach allows the UltraLight VM-UNet to maintain high segmentation performance while significantly reducing the number of parameters and computational complexity.

The UltraLight VM-UNet is reported to have only 0.049 million parameters and a computational cost of 0.060 GFLOPs, which is significantly lower than traditional convolutional neural networks and transformers used for image segmentation tasks. Despite its lightweight nature, the authors demonstrate that the UltraLight VM-UNet achieves competitive performance on three publicly available skin lesion segmentation datasets, outperforming several state-of-the-art lightweight models.

The success of the UltraLight VM-UNet can be attributed to the authors' in-depth analysis of the key factors influencing the parameters of the Vision Mamba module. By identifying the number of input channels as a critical factor affecting the parameter count, they were able to design the PVM Layer to process features in parallel while keeping the overall channel count constant, leading to a significant reduction in parameters without compromising performance.

\section{Our Models}

Inspired by \cite{U-Mamba} and \cite{ultralightvmunet}, we sought to implement both models on a different problem: tumor segmentation.

Our goal was to 1) maintain dice scores comparable to state-of-the-art (SOTA) methods such as those listed above, while 2) being smaller than those above.

We now go into what we did for each of our models.

\subsection{UMambaBot\_PPBot}

This model incorporates pyramidal pooling, which is a strategy used in convolutional neural networks (CNNs) to capture context and incorporate multi-scale information \cite{zhao2017pyramid}. Pyramidal pooling involves parallel pooling operations at different scales, followed by concatenation of the resulting feature maps. This approach has been shown to improve the performance of CNNs in various computer vision tasks, including segmentation, by enabling the model to capture both local and global context.

In our UMambaBot\_PP model, we integrated pyramidal pooling into the U-Mamba architecture, aiming to leverage the strengths of both the Mamba module and multi-scale feature extraction.

% Unsure why this comment line is giving me troubles !!

\subsection{UL-VM-UNet\_v1}
For ULMUNet\_v1, the initial channel list was modified from [8, 16, 24, 32, 48, 64] to [16, 32, 64, 128, 256], increasing the number of channels at each step of the encoder and decoder. Additionally, the depth of the U-Net was reduced by one layer to keep the parameter count low while increasing the number of parameters in each step. The intention was to improve performance by increasing the capacity of the model while maintaining a reasonable parameter count.

\subsection{UL-VM-UNet\_v2}
In ULMUNet\_v2, the number of parallel branches in the Parallel Vision Mamba (PVM) block was increased from 4 to 8. This modification directly decreases the parameter count, further proving the idea proposed in the UltraLight VM-UNet paper \cite{ultralightvmunet} that processing features in parallel with reduced channel counts can significantly reduce parameters while maintaining performance.

\subsection{UL-VM-UNet\_v3}
ULMUNet\_v3 combines the modifications from ULMUNet\_v1 and ULMUNet\_v2. It increases the number of parallel branches in the PVM block to 8 and modifies the channel list to [16, 32, 64, 128, 256]. Additionally, the depth of the U-Net was reduced to 5 layers. This approach aims to balance the trade-off between model capacity and parameter efficiency.

\subsection{UL-VM-UNet\_v4}
\gr{Pyramidal pooling was used in each of the PVM blocks in the encoder and decoder networks. If we want to do similar to UMambaBot\_PP we can do it just in the bottleneck. Let me know if we want this can I can do it fast.}
This model incorporates pyramidal pooling, similar to UMambaBot\_PP, but within the UltraLight VM-UNet architecture. The goal was to leverage the benefits of multi-scale feature extraction while maintaining the parameter efficiency of the UltraLight VM-UNet.

\subsection{UL-VM-UNet\_v5}
\gr{What is UL-PP? v5 is v4 with no PP. If any, v4 is UL-PP.}
The UL-PP (Ultra Light Pyramidal Pooling) model trained and performed well with 1.1M parameters. Incorporating pyramidal pooling on the UL network scored slightly worse than without pyramidal pooling but required approximately 400,000 more parameters.

\section{Results}
We now compare our models against ResNet18, ResNet50, standard U-Mamba bottleneck, UNet, and UltraLight VM-UNet. We train for 100 epochs on all models, and use dice-loss with Adam optimizer. We train on roughly 3000 images, and validate on roughly 400. we then test on roughly 400. We show the stats of our data in Figure \ref{fig:datastats}.
The results are shown in Figures \ref{fig:balls} and \ref{fig:segs}.

\begin{table}[ht]
    \centering
    \begin{tabular}{|l|c|c|}
        \hline
        \textbf{Model Name} & \textbf{Parameter Count (Millions)} & \textbf{Dice Score} \\
        \hline
        UMambaBot\_PP & 10.0 & 0.8867 \\
        UMambaBot & 9.8 & 0.8799 \\
        ResNet18 & 13.9 & 0.8648 \\
        ResNet50 & 42.9 & 0.8672 \\
        SegFormer & 17.8 & 0.8454 \\
        UNet & 7.8 & 0.8929 \\
        UL-VM-UNet & 0.049 & 0.79955 \\
        UL-VM-UNet\_v1 & 0.20 & 0.8409 \\
        UL-VM-UNet\_v2 & 0.042 & 0.8243 \\
        UL-VM-UNet\_v3 & 0.176 & 0.8109 \\
        UL-VM-UNet\_v4 & 1.1 & 0.8548 \\
        UL-VM-UNet\_v5 & 0.7 & 0.8633 \\
        \hline
    \end{tabular}
    \caption{Comparison of different segmentation models.}
    \label{tab:model_comparison}
\end{table}

\begin{figure}[!t]
    \centering
    \includegraphics[width=\columnwidth]{imgs/balls.png}
    \caption{A scatter plot of Dice Score vs model parameter count, where "M" means "millions. The points are scaled to help represent size.}
    \label{fig:balls}
\end{figure}

\begin{figure}[!t]
    \centering
    \includegraphics[width=\columnwidth]{imgs/masks.png}
    \caption{Example segmentations.}
    \label{fig:segs}
\end{figure}

@@

\begin{thebibliography}{00}
\bibitem{brown2020language} T. B. Brown et al., "Language Models are Few-Shot Learners," arXiv:2005.14165 [cs.CL], 2020.
\bibitem{dosovitskiy2021image} A. Dosovitskiy et al., "An Image is Worth 16x16 Words: Transformers for Image Recognition at Scale," arXiv:2010.11929 [cs.CV], 2021.
\bibitem{warden2019tinyml} P. Warden and D. Situnayake, "TinyML: Machine Learning with TensorFlow Lite on Arduino and Ultra-Low-Power Microcontrollers," O'Reilly Media, Inc., 2019.
\bibitem{banbury2021benchmarking} C. R. Banbury et al., "MLPerf Tiny Benchmark," arXiv:2106.07597 [cs.LG], 2021.
\bibitem{ronneberger2015unet} O. Ronneberger, P. Fischer, and T. Brox, "U-Net: Convolutional Networks for Biomedical Image Segmentation," arXiv:1505.04597 [cs.CV], 2015.
\bibitem{he2016deep} K. He, X. Zhang, S. Ren, and J. Sun, "Deep Residual Learning for Image Recognition," arXiv:1512.03385 [cs.CV], 2016.
\bibitem{vaswani2017attention} A. Vaswani et al., "Attention Is All You Need," arXiv:1706.03762 [cs.CL], 2017.
\bibitem{choromanski2020rethinking} K. Choromanski et al., "Rethinking Attention with Performers," arXiv:2009.14794 [cs.LG], 2020.
\bibitem{gu2022efficiently} A. Gu, K. Goel, and C. Ré, "Efficiently Modeling Long Sequences with Structured State Spaces," arXiv:2111.00396 [cs.LG], 2022.
\bibitem{gu2023mamba} A. Gu and T. Dao, "Mamba: Linear-Time Sequence Modeling with Selective State Spaces," arXiv:2312.00752 [cs.LG], 2023.
\bibitem{krizhevsky2012imagenet} - ImageNet Classification with Deep Convolutional Neural Networks
\bibitem{simonyan2014very} - Very Deep Convolutional Networks for Large-Scale Image Recognition
\bibitem{szegedy2015going} - Going Deeper with Convolutions
\bibitem{he2016deep} - Deep Residual Learning for Image Recognition
\bibitem{howard2017mobilenets} - MobileNets: Efficient Convolutional Neural Networks for Mobile Vision Applications
\bibitem{tan2019efficientnet} - EfficientNet: Rethinking Model Scaling for Convolutional Neural Networks
\bibitem{radosavovic2020designing} - Designing Network Design Spaces
\bibitem{dosovitskiy2021image} - An Image is Worth 16x16 Words: Transformers for Image Recognition at Scale
\bibitem{xie2021segformer} - SegFormer: Simple and Efficient Design for Semantic Segmentation with Transformers
\bibitem{choromanski2020rethinking} - Rethinking Attention with Performers
\bibitem{cheng2017survey} - A Survey of Model Compression and Acceleration for Deep Neural Networks
\bibitem{hinton2015distilling} - Distilling the Knowledge in a Neural Network
\bibitem{gu2022efficiently} A. Gu, K. Goel, and C. Ré, "Efficiently Modeling Long Sequences with Structured State Spaces," arXiv preprint arXiv:2111.00396, 2022.
\bibitem{gu2020hippo} A. Gu, T. Dao, S. Ermon, A. Rudra, and C. Ré, "HiPPO: Recurrent Memory with Optimal Polynomial Projections," Advances in Neural Information Processing Systems, vol. 33, pp. 1474-1487, 2020.
\bibitem{gu2023mamba} A. Gu and T. Dao, "Mamba: Linear-Time Sequence Modeling with State Spaces," arXiv preprint arXiv:2302.01327, 2023.
\bibitem{ronneberger2015unet} O. Ronneberger, P. Fischer, and T. Brox, "U-Net: Convolutional Networks for Biomedical Image Segmentation," arXiv:1505.04597 [cs.CV], 2015.

\bibitem{vaswani2023attention} A. Vaswani et al., "Attention Is All You Need," arXiv:1706.03762 [cs.CL], 2023.

\bibitem{gu2022efficiently} A. Gu, K. Goel, and C. Ré, "Efficiently Modeling Long Sequences with Structured State Spaces," arXiv:2111.00396 [cs.LG], 2022.

\bibitem{gu2023mamba} A. Gu and T. Dao, "Mamba: Linear-Time Sequence Modeling with Selective State Spaces," arXiv:2312.00752 [cs.LG], 2023.

\bibitem{bakas2019identifying} S. Bakas et al., "Identifying the Best Machine Learning Algorithms for Brain Tumor Segmentation, Progression Assessment, and Overall Survival Prediction in the BRATS Challenge," arXiv:1811.02629 [cs.CV], 2019.

\bibitem{U-Mamba} J. Ma, F. Li, and B. Wang, "U-Mamba: Enhancing Long-range Dependency for Biomedical Image Segmentation," arXiv:2401.04722 [cs.CV], 2024.

\bibitem{rombach2021highresolution} R. Rombach, A. Blattmann, D. Lorenz, P. Esser, and B. Ommer, "High-Resolution Image Synthesis with Latent Diffusion Models," arXiv:2112.10752 [cs.CV], 2021.

\bibitem{DBLP:journals/corr/abs-2006-11239} J. Ho, A. Jain, and P. Abbeel, "Denoising Diffusion Probabilistic Models," CoRR, vol. abs/2006.11239, 2020. [Online]. Available: https://arxiv.org/abs/2006.11239

\bibitem{DBLP:journals/corr/HeZRS15} K. He, X. Zhang, S. Ren, and J. Sun, "Deep Residual Learning for Image Recognition," CoRR, vol. abs/1512.03385, 2015. [Online]. Available: http://arxiv.org/abs/1512.03385

\bibitem{zhang2022segvit} B. Zhang et al., "SegViT: Semantic Segmentation with Plain Vision Transformers," arXiv:2210.05844 [cs.CV], 2022.

\bibitem{dosovitskiy2021image} A. Dosovitskiy et al., "An Image is Worth 16x16 Words: Transformers for Image Recognition at Scale," arXiv:2010.11929 [cs.CV], 2021.
\bibitem{ultralightvmunet} R. Wu, Y. Liu, P. Liang, and Q. Chang, “UltraLight VM-UNet: Parallel Vision Mamba Significantly Reduces Parameters for Skin Lesion Segmentation.” Accessed: Apr. 20, 2024. [Online]. Available: https://arxiv.org/pdf/2403.20035.pdf
\bibitem{zhao2017pyramid}
H. Zhao, J. Shi, X. Qi, X. Wang and J. Jia, "Pyramid Scene Parsing Network," 2017 IEEE Conference on Computer Vision and Pattern Recognition (CVPR), 2017, pp. 6230-6239, doi: 10.1109/CVPR.2017.660.
\bibitem{dataset} “Brain MRI segmentation,” www.kaggle.com. https://www.kaggle.com/datasets/mateuszbuda/lgg-mri-segmentation
\bibitem{xie2021segformer} E. Xie, W. Wang, Z. Yu, A. Anandkumar, J. Alvarez, and P. Luo, “SegFormer: Simple and Efficient Design for Semantic Segmentation with Transformers.” Available: https://arxiv.org/pdf/2105.15203.pdf
\end{thebibliography}
